Estimating the number of clusters using Cross Validation

### 2015-06-10 cvclust.tex
clean up elbow plots

After being proposed more than $50$ years, $K$-mean remains one of the most popular and widely used clustering algorithms \citep{jain2010data}. Like many other clustering methods, $K$-mean requires an input parameter $k$, the number of clusters, to be specified by the user. Automatically and quantitatively deciding such parameter is important and yet unsolved problem \citep{fujita2014non}. Various methods have been proposed to tackle this difficulty. One ad hoc approach is to explore the relationship between $W_k$ (within-cluster dispersion) and the number of cluster $k$ for a certain clustering method such as $K$-mean. Since $W_k$ decreases as $k$ increases, one usually find the ``elbow" of curve $W_k$ \textsl{VS} $k$ as the appropriate number of clusters. The example on the top row of Figure \ref{fig1} demonstrates such approach for data with $k=4$, where the ``elbow" point indeed reveals the true number of clusters. This is based on the idea that under partitioning data set has more impact than over partitioning data set in terms of $W_k$. However, locating the ``elbow" point is somewhat subjective and sometimes is not appropriate to select the optimal $k$. The second example on the bottom row of Figure \ref{fig1} shows a  situation where there is no clear choice of the ``elbow" point -- both $k=2$ and $k=3$ can be viewed as the ``elbow" point. What's more, the true $k=4$ can never be selected as the optimal $k$ using such approach in this case since it can hardly be viewed as the ``elbow" of the curve.
\begin{figure}
\begin{minipage}{\linewidth}
\centering
\begin{minipage}{0.45\linewidth}
\includegraphics[width=\linewidth]{figure/Data_example.pdf}
\end{minipage}
\hspace{0.05in}
\begin{minipage}{0.45\linewidth}
\includegraphics[width=\linewidth]{figure/W_k.pdf}
\end{minipage}
\end{minipage}
%\center{\caption{\label{fig1} Left plot shows the actual data, and the right plot shows the $W_k$ \textsl{VS} $k$ curve. }}
%\end{figure}

%\begin{figure}
\begin{minipage}{\linewidth}
\centering
\begin{minipage}{0.45\linewidth}
\includegraphics[width=\linewidth]{figure/Data_elbow.pdf}
\end{minipage}
\hspace{0.05in}
\begin{minipage}{0.45\linewidth}
\includegraphics[width=\linewidth]{figure/elbow.pdf}
\end{minipage}
\end{minipage}
\center{\caption{\label{fig1} Left plots show the actual data, right plots show the corresponding $W_k$ \textsl{VS} $k$ curve. }}
\end{figure}
\\

>>>>

After being proposed more than $50$ years, $K$-mean remains one of the most
popular and widely used clustering algorithms \citep{jain2010data}. Like many
other clustering methods, $K$-mean requires an input parameter $k$, the number
of clusters, to be specified by the user. Automatically and quantitatively
deciding such parameter is important and yet unsolved problem
\citep{fujita2014non}. Various methods have been proposed to tackle this
difficulty. One ad hoc approach is to explore the relationship between $W_k$
(within-cluster dispersion) and the number of cluster $k$ for a certain
clustering method such as $K$-mean. Since $W_k$ decreases as $k$ increases,
one usually find the ``elbow" of curve $W_k$ \textsl{VS} $k$ as the
appropriate number of clusters. The example on the top row of Figure
\ref{fig:elbow} demonstrates such approach for data with $k=4$, where the
``elbow" point indeed reveals the true number of clusters. This is based on
the idea that under partitioning data set has more impact than over
partitioning data set in terms of $W_k$. However, locating the ``elbow" point
is somewhat subjective and sometimes is not appropriate to select the optimal
$k$. The second example on the bottom row of Figure \ref{fig:elbow} shows a  situation where there is no clear choice of the ``elbow" point -- both $k=2$ and $k=3$ can be viewed as the ``elbow" point. What's more, the true $k=4$ can never be selected as the optimal $k$ using such approach in this case since it can hardly be viewed as the ``elbow" of the curve.

\begin{figure}
\centering
\begin{minipage}{\linewidth}
  \begin{minipage}{0.45\linewidth}
    \includegraphics[width=\linewidth]{figure/elbow-correct-data.pdf}
  \end{minipage}
  \hspace{0.05in}
  \begin{minipage}{0.45\linewidth}
    \includegraphics[width=\linewidth]{figure/elbow-correct-withinss.pdf}
  \end{minipage}
\end{minipage}
\begin{minipage}{\linewidth}
  \begin{minipage}{0.45\linewidth}
    \includegraphics[width=\linewidth]{figure/elbow-incorrect-data.pdf}
  \end{minipage}
  \hspace{0.05in}
  \begin{minipage}{0.45\linewidth}
    \includegraphics[width=\linewidth]{figure/elbow-incorrect-withinss.pdf}
  \end{minipage}
\end{minipage}
\caption{Left panels show the $(X,Y)$ data points; right panels
  show the corresponding values of the within-cluster sum of squares $W_k$
plotted against the number of clusters, $k$.}
\label{fig:elbow}
\end{figure}
